---
title: "Peer Assignment-1"
author: "Socrates"
date: "Saturday, February 14, 2015"
output: html_document
---

<br>
<br>

### 1. Loading and preprocessing the data
#### Setting Working Directory 
```{r, echo=TRUE}
source_url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
target_file <- "activity.zip"
dest_folder <- "C:\\courses\\5-Reproducible Research\\Project"
data_folder <- "C:\\courses\\5-Reproducible Research\\Project\\data"
setwd(dest_folder)
```

#### Downloading files
```{r, echo=TRUE}
setInternet2(use = TRUE)
download.file(source_url, target_file)
unzip(target_file, exdir=data_folder)
```

#### Loading Data into R working Enviornment
```{r, echo=TRUE}
file <- list.files(data_folder, full.names=TRUE)
data <- read.csv(file, sep=",", header=TRUE, colClasses=c('numeric', 'character', 'numeric'))
```
<br>
<br>

### 2. What is mean total number of steps taken per day?

#### Convert Date class from <code>Character</code> to <code>Date</code>
```{r, echo=TRUE}
data$date <- as.Date(data$date, format="%Y-%m-%d")
```

#### Calculate the total number of steps taken per day
```{r, echo=TRUE}
totalval <- aggregate(steps ~ date, data = data, sum)
```

```{r, echo=FALSE}
origsum <- sum(totalval[,2])
origcnt <- length(totalval[,2])
```

#### Make a histogram of the total number of steps taken each day
```{r, echo=TRUE}
hist(totalval[,2], xlab="No of Steps", main='', breaks = 30, col='blue')
```

#### Calculate Mean value
```{r, echo=TRUE}
meanval <- round(mean(totalval[,2], na.rm=TRUE))
meanval
```

#### Calculate Median value
```{r, echo=TRUE}
medianval <- round(median(totalval[,2], na.rm=TRUE))
medianval
```
<br>
<br>

### 3. What is the average daily activity pattern?

#### Time series plot. x-axis: Average number of steps taken; y-axis: Averaged across all days
```{r, echo=TRUE}
meaninterval <- aggregate(steps ~ interval, data=data, mean, na.rm=TRUE)
plot(meaninterval$steps ~ meaninterval$interval, type = "l", xlab='Period', col='blue', ylab='Average No of Steps', main='')
```

#### Which 5-minute interval containing maximum number of steps?
```{r, echo=TRUE}
meaninterval[(meaninterval$steps==max(meaninterval$steps)),1]
```
<br>
<br>

### 4. Imputing missing values
#### Calculate and report the total number of missing values in the dataset
```{r, echo=TRUE}
nrow(data[data$steps == 'NA',])
```

#### Strategy for filling in all of the missing values: The mean for that 5-minute interval.
```{r, echo=TRUE}
meaninterval <- aggregate(steps ~ interval, data=data, mean, na.rm=TRUE)
meaninterval$steps <- round(meaninterval$steps)
```


#### Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo=TRUE}
for (i in 1:nrow(meaninterval)) {

  # returns 0, 5, 10, 15, ...
	intr.val <- meaninterval$interval[i]
	upd.df <- length(data[which(data$interval == intr.val & is.na(data$steps)),]$steps)
	if (upd.df > 0) {
		data[which(data$interval == intr.val & is.na(data$steps)),]$steps <- meaninterval$steps[meaninterval$interval == intr.val]
	}
}
```


#### Make a histogram of the total number of steps taken each day on new dataset
```{r, echo=TRUE}
totalval <- aggregate(steps ~ date, data = data, sum)
hist(totalval[,2], xlab="No of Steps", main='', col='blue', breaks = 30)
```


```{r, echo=FALSE}
revsum <- sum(totalval[,2])
revcnt <- length(totalval[,2])
```

#### Calculate and report Mean total number of steps taken per day.
```{r, echo=TRUE}
meanval <- round(mean(totalval[,2], na.rm=TRUE))
meanval
```

#### Calculate and report Median total number of steps taken per day.
```{r, echo=TRUE}
medianval <- round(median(totalval[,2], na.rm=TRUE))
medianval
```

#### Do these values differ from the estimates from the first part of the assignment? 
<code>YES, They differ.</code>

#### What is the impact of imputing missing data on the estimates of the total daily number of steps?
<code>
After imputing missing data, the Total Sum got increased from `r origsum` to `r revsum`. <br>
However the mean and median values remain same. This is due to the fact that the no of days <br>
in original dataset increased from `r origcnt` to `r revcnt` after populating values for "NA"</code>
<br>
<br>



### 5. Are there differences in activity patterns between weekdays and weekends?

#### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" 
```{r, echo=TRUE}
data$dayname <- weekdays(data$date)
data$daytype <- ifelse(data$dayname %in% c("Saturday", "Sunday"), "weekend", "weekday")
```

#### Panel plot containing a time series plot 
```{r, echo=TRUE}
library(lattice) 
meansteps <- aggregate(data$steps, by = list(data$interval, data$daytype), mean)
names(meansteps) <- c("interval", "daytype", "steps")
xyplot(steps ~ interval | daytype, meansteps, type = "l", layout = c(1, 2), xlab = "Interval", ylab = "Number of steps")
```